Classless Java: Tuning Java Interfaces

### 2015-12-02 paper/sections/CaseStudy.tex
revise on related work. TODO: ‘one method limitation discussion’

\paragraph{Adding a New Type of Expressions}
In the object oriented paradigm, it is easy to add new types of expressions. For example, the following code shows how to add subtraction.

@@

\marco{text under need to be improved}
\paragraph{Adding a New Operation} The difficulty of the EP in object-oriented 
languages arises from adding new operations. For example, adding a pretty printing 
operation, would typically involve changing all of the existing code. However, a solution 
to the EP forbids this and it also forbids using casts. In other words, it should be possible 
to add the pretty printing operation in a type-safe and modular way.  
It turns out that, this can be done easily, using our object interfaces approach. 
The code in Figure~\ref{fig:ep} shows how to add the new operation \texttt{print}.

\noindent  The interface \texttt{ExpP} extending interface \texttt{Exp} is defined with the
extra method \texttt{print()}. Interfaces \texttt{LitP} and \texttt{AddP} are
defined with default implementations of \texttt{print()}, extending
base interfaces \texttt{Lit} and \texttt{Add}, respectively. Importantly, note
that in \texttt{AddP}, the types of the ``\emph{fields}'' (i.e. the getter methods) \texttt{e1} 
and \texttt{e2} are refined. If the types were not refined then the body of the \texttt{print}
method in \texttt{AddP} would fail to type-check.

>>>>

\paragraph{Adding a New Type of Expressions}
In the OO paradigm, it is easy to add new types of expressions. For example, the
following code shows how to add subtraction.

@@

%\marco{text under need to be improved}
\paragraph{Adding a New Operation} The difficulty of the EP in OO  
languages arises from adding new operations. For example, adding a pretty printing 
operation would typically change all existing code. However, a solution 
%to the EP forbids this and it also forbids using casts. In other words, it should be possible 
should add the pretty printing operation in a type-safe and modular way.  It
turns out to be easily achieved with the assitance of \mixin.  The code in
Figure~\ref{fig:ep} (rhs) shows how to add the new operation \texttt{print}.
Interface \texttt{ExpP} extends \texttt{Exp} with the extra method
\texttt{print()}. Interfaces \texttt{LitP} and \texttt{AddP} are defined with
default implementations of \texttt{print()}, extending base interfaces
\texttt{Lit} and \texttt{Add}, respectively. Importantly, note that in
\texttt{AddP}, the types of ``\emph{fields}'' (i.e. the getter methods)
\texttt{e1} and \texttt{e2} are refined. If the types were not refined then
\texttt{print()} method in \texttt{AddP} would fail to type-check.

### 2015-12-02 paper/sections/RelatedWork.tex
fix some spacing issues
revising text
revise on related work. TODO: ‘one method limitation discussion’

\subsection{Multiple Inheritance in Object Oriented Languages}
Many authors have argued in favour or against multiple inheritance.  Multiple inheritance provides
expressive power, but it is difficult to model and implement, and can create
programs that are hard to reason about. These difficulties include the famous
diamond (fork-join) problem~\cite{bracha90mixin,Sak89dis}, conflicting methods, etc.
%and the yo-yo problem~\cite{taenzer1989problems}.  
To conciliate the need for expressive power and the need for simplicity, many
models have been proposed in over the years, including C++ virtual
inheritance, mixins~\cite{bracha90mixin}, traits~\cite{scharli03traits}, and
hybrid model such as CZ~\cite{malayeri2009cz}.
They provide novel programming architecture models in the OO paradigm. In terms
of restrictions set on these models, C++ virtual inheritance aims at a more
general model, mixins added some restrictions on the model, and trait model is
the most restricted one (excluding states, instantiation, etc).

\paragraph{C++ Approach.}
C++ tries to provide a general solution to multiple inheritance. Virtual
inheritance in C++ provides another solution to multiple inheritance (especially
the diamond problem by keeping only one copy of the base
class)~\cite{ellis1990annotated}, however suffers from object initialization
problem as pointed out by Malayeri et al.~\cite{malayeri2009cz}. It bypasses all
constructor calls to virtual superclasses, which would potentially cause serious
semantic errors. In our approach, \mixin annotation processor has the full
control of object initialization mechanism, and this mechanism is clean and
transparent to users. Moreover, if not satisfied with the default generated
\texttt{of} method, customized factory methods are also allowed to be defined.

\paragraph{Mixins.}
Mixins are a more restricted model than the C++ approach. Mixins allow to name
components that can be applied to various classes as reusable functionality
units. However, they suffer from linearisation: the order of mixin application
is relevant in often subtle and undesired ways. This hinders their usability
and the ability of resolving conflicts: the linearisation (total ordering) of
mixin inheritance cannot provide a satisfactory resolution in all cases and
restricts the flexibility of mixin composition. To fight those limitations, an
algebra of mixin operators is introduced~\cite{ancona2002calculus}, but this
raised the complexity of the approach, especially when constructors and fields
are considered~\cite{marco09FJigsaw}. Scala traits are in fact more like linearised mixins.
Scala avoids the object initialization
problem by disallowing constructor parameters, causing no ambiguity in cases
such as diamond problem. However this approach has limited expressiveness, and
suffers from all the problems of linearised mixin composition. Other languages, such as 
Python, also use linearized mixins.
\begin{comment}
Python also offers multiple inheritance via linearised mixins. Indeed in python any class is implicitly a mixin, and mixin composition informally expressed as\\*
\Q@ class A use B,C {...new methods...}@\\*
can be expressed in python as \\*
\Q@ class Aux: ...new methods...@\\*
\Q@ class A(B,C,Aux): pass@ 
\end{comment}
\noindent Java interfaces and default methods does not use
linearisation: the semantics of Java \textbf{extends} clause in
interfaces is unordered and symmetric.

@@

\paragraph{CZ.}
Malayeri and Aldrich proposed a model CZ~\cite{malayeri2009cz} which aims to do
multiple inheritance without diamond problem. They divide inheritance into two
separate concepts: inheritance dependency (using \texttt{require}) and
implementation inheritance (using \texttt{extends}). Using a combination of
\texttt{requires} and \texttt{extends}, a program with diamond inheritance can be
transformed to one without diamonds. Also in this approach, fields and multiple
inheritance can coexist. However as shown in the example in the article, the
untangling of inheritance also untangles class structure. In CZ, not only the
number of classes (from 4 to 6 in the stream example), but also the class
hirarchy complexity increases. Our approach does not complicate the structure,
and (conceptual) fields coexist with multiple inheritance.

\paragraph{Traits and Java's default methods}
Simplifying the mixins approach, traits~\cite{scharli03traits} draw a
strong line between units of reuse (traits) and object factories
(classes). In this model, traits as units of reusable code, contain
only methods as reusable entities. Thus, no state and state
initialization are considered. Classes act as object factories,
requiring functionalities from multiple traits. Java 8 interfaces with
default methods are closely related to traits: concrete method
implementation are allowed (via the \textbf{default} keyword) inside
interfaces. The introduction of default methods opens the gate for
various flavours of multiple inheritance in Java, using interfaces.
Traits offer a trait algebra with operations like sum, alias and exclusion, provided
for explicit conflict resolution.  
Former work by Bono et al.~\cite{bono14}. provides details on
mimicking the trait algebra through Java 8 interfaces. 
We briefly recall the main points of their encoding;
however we propose a different representation of \textbf{exclusion}.
The author of ~\cite{bono14} agree that our revised version is
cleaner, typesafe and more direct.
%\bruno{do we want do say this?}\yanlin{need to ask Marco.}

\newcommand\shortItem{\vspace{-1ex}}
\begin{itemize}
\item \textbf{Symmetric sum} can be obtained by simple multiple inheritance between interfaces.
\shortItem
    \begin{lstlisting}
    interface A { int x(); }    interface B { int y(); }    interface C extends A, B {}
    \end{lstlisting}
\shortItem
\item \textbf{Overriding} a conflict is obtained by specifying which super interface take precedence.
\shortItem
    \begin{lstlisting}
    interface A { default int m() {return 1;} } 
    interface B { default int m() {return 2;} }
    interface C extends A, B { default int m() {return B.super.m();} }
    \end{lstlisting}
\shortItem
\item \textbf{Alias} is creating  a new method delegating to the existing super interface.
\shortItem
    \begin{lstlisting}
    interface A { default int m() {return 1;} }
    interface B extends A { default int k() {return A.super.m();} }
    \end{lstlisting}
\shortItem

\item \textbf{Exclusion}: exclusion is also supported in Java, where method declarations can hide the default methods correspondingly in the super interfaces.
\shortItem
    \begin{lstlisting}
    interface A { default int m() {return 1;} }
    interface B extends A { int m(); }
    \end{lstlisting}
\shortItem
\end{itemize}

There are also proposals for extending Java (before Java8) with traits. For
example, FeatherTrait Java (FTJ)~\cite{Liquori08ftj} by Liquori et al. extends
the calculus of Featherweight Java (FJ)~\cite{Igarashi01FJ} with
statically-typed traits, adding trait-based inheritance in Java.  Except for
few, mostly syntactic details, their work can be emulated/rephrased with Java8
interfaces.  There are also extensions to the original trait model, which have
operations that default methods and interfaces cannot model, such as method
renaming in \cite{reppy2006foundation}, which breaks structural subtyping.

\paragraph{Traits vs Object Interfaces}
We consider object interfaces to be an alternative to traits or mixins.  In the
trait model two concepts (traits and classes) coexist and cooperate. Some
authors~\cite{BettiniDSS13} see this as good language design fostering good
software development by helping programmers to think about the structure of
their programs.  However, other authors see the need of two concepts and the
absence of state as drawbacks of this model~\cite{malayeri2009cz}.  Object
interfaces are units of reuse, and at the same time they provide factory methods
for instantiation and support state.  Our approach promotes the use of
interfaces instead of classes, in order to rely on the modular composition
offered by interfaces. Since Java was designed for classes, a direct class-less
programming style is verbose and feels unnatural. However, annotation driven
code generation is enough to overcome this difficulty, and the resulting
programming style encourages modularity, composability and reusability, by
keeping a strong object oriented feel. In that sense, we promote object
interfaces as being both units of reusable code and object factories. Our
practical experience is that, in Java, separating the two notions leads to alot
of boilerplate code, and is quite limiting when multiple inheritance with state
is required. The use of abstract state operations avoids the key difficulties
associated with multiple inheritance and state, while still being quite
expressive.  Moreover the ability to support constructors adds additional
expressivity, which is not available in approaches such as Scala's
traits/mixins.

@@

\subsection{Meta-programming Competes with Language Extensions}
\bruno{needs to be polished better, I think}
\bruno{Mention/Compare with Language Virtualization work in Scala. They also re-interpret 
syntax.}
The most obvious solution to adding features to a language is language
extension.  It is often implemented as syntactic extensions that can be
desugared to the base language. For example, the Scala compiler was extended to
directly supports XML syntax. However, this approach does not support combining
multiple extensions into one. We are de facto creating a fork in the language,
and rarely the new fork gain enough traction to become the main language
release.  On this topic we mention SugarJ~\cite{erdweg11sugarj} - a Java-based
extensible language allowing programmers to extend it with custom features by
definitions in meta-DSLs (SDF, Stratego, etc).\bruno{so, what's the point that 
we want to make with SugarJ?}\yanlin{we were talking about lightweight language
sugar, so SugarJ is a comparison to ours. Where using SugarJ requires learning
various DSLs like SDF, stratego, etc, in comparison, our approach is just using
Java. }

>>>>

\subsection{Multiple Inheritance in Object Oriented Languages}
Many authors have argued in favour or against multiple inheritance.  Multiple
inheritance is very expressive, but difficult to model and implement, and can
cause difficulty (including the famous diamond (fork-join)
problem~\cite{bracha90mixin,Sak89dis}, conflicting methods, etc.) in reasoning
programs. To conciliate the need for expressive power and simplicity, many
models have been proposed, including C++ virtual inheritance,
mixins~\cite{bracha90mixin}, traits~\cite{scharli03traits}, and hybrid model
such as CZ~\cite{malayeri2009cz}.  They provide novel programming architecture
models in the OO paradigm. In terms of restrictions set on these models, C++
virtual inheritance aims at a relative general model; mixin model add some
restrictions on the model; and trait model is the most restricted one (excluding
states, instantiation, etc).

\paragraph{C++ Approach.}
C++ tries to provide a general solution to multiple inheritance by virtual
inheritance that deals with diamond problem by keeping only one copy of the base
class~\cite{ellis1990annotated}, however suffers from object initialization
problem as shown in~\cite{malayeri2009cz}. It bypasses all constructor calls to
virtual superclasses, which would potentially cause serious semantic errors. In
our approach, \mixin annotation processor has the full control over object
initialization, whose mechanism is transparent to users. Moreover, customized
factory methods are also allowed, if users are not satisfied with the default
generated \texttt{of} method.

\paragraph{Mixins.}
Mixins are a more restricted model than C++ approach. Mixins allow to name
components that can be applied to various classes as reusable functionality
units. However, % they suffer from linearisation: the order of mixin application
% is relevant in often subtle and undesired ways. This hinders their usability
% and the ability of resolving conflicts:
the linearisation (total ordering) of
mixin inheritance cannot provide a satisfactory resolution in some cases and
restricts the flexibility of mixin composition. To fight this limitation, an
algebra of mixin operators is introduced~\cite{ancona2002calculus}, but this
raises the complexity, especially when constructors and fields
are considered~\cite{marco09FJigsaw}. Scala traits are in fact more like linearised mixins.
Scala avoids the object initialization
problem by disallowing constructor parameters, causing no ambiguity in cases
such as diamond problem. However this approach has limited expressiveness, and
suffers from all the problems of linearised mixin composition. Other languages, such as 
Python, also use linearized mixins.
\begin{comment}
Python also offers multiple inheritance via linearised mixins. Indeed in python any class is implicitly a mixin, and mixin composition informally expressed as\\*
\Q@ class A use B,C {...new methods...}@\\*
can be expressed in python as \\*
\Q@ class Aux: ...new methods...@\\*
\Q@ class A(B,C,Aux): pass@ 
\end{comment}
\noindent Java interfaces and default methods does not use
linearisation: the semantics of Java \textbf{extends} clause in
interfaces is unordered and symmetric.

@@

\paragraph{CZ.}
Malayeri and Aldrich proposed a model CZ~\cite{malayeri2009cz} which aims to do
multiple inheritance without diamond problem. They divide inheritance into two
separate concepts: inheritance dependency and implementation inheritance. Using
a combination of \texttt{requires} and \texttt{extends}, a program with diamond
inheritance can be transformed to one without diamonds. Moreover, fields and
multiple inheritance can coexist. However as shown in the example in the
article, the untangling of inheritance also untangles class structure. In CZ,
not only the number of classes (from 4 to 6 in the stream example), but also the
class hirarchy complexity increases. \mixin does not complicate the
structure, and fields can also coexist with multiple inheritance.

\paragraph{Traits and Java's default methods}
Simplifying the mixins approach, traits~\cite{scharli03traits} draw a strong
line between units of reuse and object factories. Traits, as units of reusable
code, contain only methods as reusable functionality. ignoring states and state
initialization. Classes, as object factories, require functionalities from
(multiple) traits. Java 8 interfaces with default methods are closely related to
traits: concrete method implementation are allowed (via the \textbf{default}
keyword) inside interfaces. The introduction of default methods opens the gate
for various flavours of multiple inheritance in Java. Traits offer an algebra
with operations like sum, alias and exclusion, providing explicit conflict
resolution. Former work~\cite{bono14} provides details on mimicking the trait
algebra through Java 8 interfaces.  We briefly recall the main points of their
encoding; however we propose a different representation of \textbf{exclusion}.
The author of ~\cite{bono14} agrees that our revised version is cleaner,
typesafe and more direct.

\newcommand\shortItem{\vspace{-1ex}}
\begin{itemize}
\item \textbf{Symmetric sum} can be obtained by simple multiple inheritance between interfaces.
    \begin{lstlisting}
    interface A { int x(); }    interface B { int y(); }    interface C extends A, B {}
    \end{lstlisting}
\shortItem
\item \textbf{Overriding} a conflict is obtained by specifying which super interface take precedence.
    \begin{lstlisting}
    interface A { default int m() {return 1;} } 
    interface B { default int m() {return 2;} }
    interface C extends A, B { default int m() {return B.super.m();} }
    \end{lstlisting}
\shortItem
\item \textbf{Alias} is creating  a new method delegating to the existing super interface.
    \begin{lstlisting}
    interface A { default int m() {return 1;} }
    interface B extends A { default int k() {return A.super.m();} }
    \end{lstlisting}
\shortItem

\item \textbf{Exclusion}: exclusion is also supported in Java, where method declarations can hide the default methods correspondingly in the super interfaces.
    \begin{lstlisting}
    interface A { default int m() {return 1;} }
    interface B extends A { int m(); }
    \end{lstlisting}
\shortItem
\end{itemize}

There are also proposals for extending Java (before Java8) with traits. For
example, FeatherTrait Java (FTJ)~\cite{Liquori08ftj} extends
FJ~\cite{Igarashi01FJ} with statically-typed traits, adding trait-based
inheritance in Java.  Except for few, mostly syntactic details, their work can
be emulated with Java8 interfaces. There are also extensions to the original
trait model, which have operations that default methods and interfaces cannot
model, such as method renaming in \cite{reppy2006foundation}, which breaks
structural subtyping.

\paragraph{Traits vs Object Interfaces}
We consider object interfaces to be an alternative to traits or mixins.  In the
trait model two concepts (traits and classes) coexist and cooperate. Some
authors~\cite{BettiniDSS13} see this as good language design fostering good
software development by helping programmers to think about the structure of
their programs.  However, other authors see the need of two concepts and the
absence of state as drawbacks of this model~\cite{malayeri2009cz}.  Object
interfaces are units of reuse, and at the same time they provide factory methods
for instantiation and support state.  Our approach promotes the use of
interfaces instead of classes, in order to rely on the modular composition
offered by interfaces. Since Java was designed for classes, a direct classless
programming style is verbose and unnatural. However, annotation driven code
generation is enough to overcome this difficulty, and the resulting programming
style encourages modularity, composability and reusability, by keeping a strong
OO feel. In that sense, we promote object interfaces as being both units of
reusable code and object factories. Our practical experience is that, in Java,
separating the two notions leads to a lot of boilerplate code, and is quite
limiting when multiple inheritance with states is required.  Abstract state
operations avoid the key difficulties associated with multiple inheritance and
states, while still being quite expressive.  Moreover the ability to support
constructors adds additional expressivity, which is not available in approaches
such as Scala's traits/mixins.

@@

\subsection{Meta-programming Competes with Language Extensions}
The most obvious solution to adding features to a language is language
extension. It is often implemented as syntactic extensions that can be desugared
to the base language. For example, the Scala compiler was extended to directly
supports XML syntax. However, this approach does not support combining multiple
extensions into one. Scala-Virtualized is an extension to Scala which allows to
blend shallow and deep embedding of DSLs. It redefines Scala structures to
method calls, which can be overridden by DSL implementors. Therefore
Scala-Virtualized also reinterprets syntax. These extensions are admittedly
useful, however they are essentially creating a fork of the language, and rarely
the new fork gains enough traction to become the main language release. % On this
% topic we mention
SugarJ~\cite{erdweg11sugarj} is a Java-based extensible language allowing
programmers to extend it with custom features by definitions in meta-DSLs (SDF,
Stratego, etc). Using SugarJ requires learning various DSLs while our approach
is just using Java. However its intention is different from ours: one of the
purpose of SugarJ is to allow easy defition of sugars while our approach aims to
implement the desired program transformation without just in Java.
